---
title: "Carol Flynn's PhD Roots Metabarcoding Data Architecture"
author: "Carol Flynn", "Gabriel A. Mendes"
format: html
editor: visual
editor_options:
  chunk_output_type: console
chunk_output_type: console
---

# Mock Communities ASV versus MOTUs project - ASV analyses

## 1. Setup & Configuration

### Packages

-   **Load libraries.** Import all necessary packages for data manipulation, demultiplex, ASVs inference, taxonomy assigment and data visualization.

    ```{r}
    library(tidyverse)
    library(phyloseq)
    library(Biostrings)
    library(ShortRead)
    library(dada2)
    library(DECIPHER)
    library(future)
    library(viridis)
    library(ggh4x)
    library(furrr)
    library(writexl)
    library(tictoc)
    library(BLASTr)
    library(janitor)
    library(readxl)
    ```

### Directory Management
-   **Create directory structure.** Define project paths and create the directory structure required by the pipeline.

    ```{r}
    # Project folder path  
    prjct_path <- "/home/gabriel/projetos/roots_metabar"

    # Derived paths
    {
      env_path <- file.path(prjct_path, "environments")
      data_path <- file.path(prjct_path, "data")
      pipe_libs <- file.path(data_path, "reads")
      raw_libs <- file.path(pipe_libs, "raw")
      paired_libs <- file.path(pipe_libs, "paired")
      cutadapt_libs <- file.path(pipe_libs, "cutadapt")
      qual_path <- file.path(prjct_path, "quality")
      results_path <- file.path(prjct_path, "results")
      figs_path <- file.path(prjct_path, "figures")
      blast_path <- file.path(results_path, "blast")
      swarm_path <- file.path(results_path, "swarm")
      dmx_path <- file.path(pipe_libs, "demultiplex")
      nochim_path <- file.path(results_path, "non_chimera")
      tables_path <- file.path(results_path, "tables")
      track_path <- file.path(results_path, "track_plots")
      error_path <- file.path(results_path, "error_plot")
    }

    # Paths to be created
    paths <- c(prjct_path, env_path, data_path, pipe_libs, raw_libs,
               paired_libs, cutadapt_libs, qual_path, results_path,
               figs_path, blast_path, swarm_path, dmx_path, nochim_path, 
               tables_path, track_path, error_path)

    for (dir in paths) {
      if (dir.exists(dir)) {
    print(paste("The folder", dir, "already exists!"))
      } else {
    print(paste("Creating", dir, "!" ))
    dir.create(dir) 
      }
    }

    # Files in the project folder
    list.files(prjct_path)
    ```
    
### Raw data aquisition

-   **Upload raw metadata sheets**

    ```{r}
    # ITS Sequencing Metadata from Metagen
    ITS_sequecing_data_tbl <- 
      read_csv(paste0(data_path, "/", "raw/",
                      "infos_carol_its.csv")) %>%
      tibble()
    
    # RbcL Sequencing Metadata from Metagen
    RbcL_sequecing_data_tbl <- 
      read_csv(paste0(data_path, "/", "raw/",
                      "infos_carol_rbcl.csv")) %>%
      tibble()
    
    # Sample Metadata 
    smpl_metadata_tbl <- 
      read_csv(paste0(data_path,"/", "raw/",
                      "soil_complete_2026.csv")) %>%
      tibble()
    
    # Soil Core Metadata 
    soil_metadata_tbl <- 
      read_csv(paste0(data_path,"/", "raw/",
                      "soilcoredata_flynn_2026.csv")) %>%
      tibble()
    
    # Associated Core Species Data
    species_per_core_tbl <- 
      read_excel(paste0(data_path,"/", "raw/",
                        "species_per_core_20m.xlsx")) %>%
      tibble()
    ```

## 2. Raw Data Pre-processing

-   **Formatting Metadata Tables**

    ```{r}
    # Merging and Correcting Sequencing Metadata table
    sequencing_data_tbl <-
      rbind(ITS_sequecing_data_tbl,
            RbcL_sequecing_data_tbl) %>%
      mutate(Lab_ID = case_when(!str_detect(sample, "^CN") ~ 
                                  paste0("L", str_extract(sample, "^[^.]+")),
                                TRUE ~ NA),
             sample = case_when(!str_detect(sample, "^CN") ~ 
                                  paste0("L", sample),
                                TRUE ~ paste0(CN, "_", pool)),
             replica = case_when(!str_detect(sample, "^CN") ~ 
                                  str_extract(sample, "[^.]+$"),
                                TRUE ~ NA),
             replica = case_when(replica == "1" ~ "A",
                                 replica == "2" ~ "B",
                                 replica == "3" ~ "C",
                                 TRUE ~ replica),
             obs = paste(forward_tag_name, reverse_tag_name, sep = "_")) %>%
      relocate(Lab_ID, pool, sample, replica)
    
    # Merging Sample and Soil Core Metadata table
    sample_metadata_tbl <-
    smpl_metadata_tbl %>% 
      mutate(Lab_ID = paste0(Core_ID, Depth_m)) %>% # Fixing the Lab_ID values 
      select(-c(Core_ID, Depth_m)) %>% 
      left_join(soil_metadata_tbl, by = "Lab_ID") %>% 
      relocate(Lab_ID)
    
    # Formatting Associated Core Species data table
    species_per_core_tbl_rdx <-
      species_per_core_tbl %>%
      mutate(Genus = word(species_name, 1),
             Species = word(species_name, 1, 2)) %>%
      group_by(Core_ID) %>%
      summarise(Core_Families = paste(sort(unique(familia)), collapse = " | "),
                Core_Genera  = paste(sort(unique(Genus)), collapse = " | "),
                Core_Species = paste(sort(unique(Species)), collapse = " | "),
                Core_Abundance = paste(sort(unique(abundance)), collapse = " | "),
                .groups = "drop")
    ```

-   **Merging Metadata Tables**

    ```{r}
    # Merging 
    primers_n_samples <-
      sequencing_data_tbl %>% 
      left_join(sample_metadata_tbl, by = "Lab_ID") %>% 
      left_join(species_per_core_tbl_rdx, by = "Core_ID") %>% 
      relocate(Core_ID, Lab_ID, sample, replica) %>% 
      mutate(`Researcher` = "Carol Flynn",
             `Run` = "mtg_jun26",
             `Project` = "roots_metabar",
             `Primer` = str_extract(forward_tag_name, "^[^-]*"),
             `Type` = case_when(str_detect(sample, "^CN") ~ "PCR Control", 
                                TRUE ~ "Sample"),
             `Metadata 2` = Lab_ID,
             lib = sample,
             `Extraction control` = NA,
             `PCR control` = case_when(!str_detect(sample, "^CN") ~ CN, 
                                TRUE ~ NA),
             `Filtration control` = NA,
             Lib = sample,
             `Forward_sequence` = paste(forward_tag, forward_primer),
             `Reverse_sequence`= paste(reverse_tag, reverse_primer),
             `Multiplexed data path` = "/home/gabriel/projetos/roots_metabar/data/reads/raw/multiplexed",
             `Demultiplexed data path` = "/home/gabriel/projetos/roots_metabar/data/reads/demultiplex") %>% 
      rename(Lab_ID = "Sample",
             sample = "Unique_File_name",
             Core_ID = "Metadata 1",
             replica = "Metadata 3",
             pool = "Metadata 4",
             `pH-H2O` = "Metadata 5",
             P = "Metadata 6",
             K = "Metadata 7",
             `Ca2+` = "Metadata 8",
             `Mg2+` = "Metadata 9",
             `Al3+` = "Metadata 10",
             MO = "Metadata 11",
             `P-Rem` = "Metadata 12",
             `mean_height_5m` = "Metadata 13",
             `max_height_5m` = "Metadata 14",
             `mean_height_10m` = "Metadata 15",
             `max_height_10m` = "Metadata 16",
             `n_bigtrees_20m` = "Metadata 17",
             `min_dist_big_m` = "Metadata 18",
             Date = "Metadata 19",
             Line = "Metadata 20",
             Transect = "Metadata 21",
             Zone = "Metadata 22",
             `Depth_m` = "Metadata 23",
             TransectGroup = "Metadata 24",
             `Cumulative_Root_g` = "Metadata 25",
             `Gravimetric_WC` = "Metadata 26",
             `Root_Density_g_m3` = "Metadata 27",
             `Core_Families` = "Metadata 28",
             `Core_Genera` = "Metadata 29",
             `Core_Species` = "Metadata 30",
             `Core_Abundance` = "Metadata 31",
             `forward_tag_name` = "Index name FWD",
             `Forward_sequence` = "Sequence FWD",
             `reverse_tag_name` = "Index name REV",
             `Reverse_sequence` = "Sequence REV") %>% 
      select(-c(CN)) %>% 
      relocate(Researcher, Run, Project, Primer, Sample, Unique_File_name, Type,
               num_range("Metadata ", 1:31), obs, `Extraction control`, `PCR control`, 
               `Filtration control`, Lib, `Index name FWD`, `Sequence FWD`,
               `Index name REV`, `Sequence REV`, `Multiplexed data path`,
               `Demultiplexed data path`)
    
    
    # Saving
    write.csv(primers_n_samples,
      file = paste0(data_path, paste0("/roots_metabar-primers_n_samples-", Sys.Date(), ".csv")))
    ```

