---
title: "Carol Flynn's PhD Roots Metabarcoding: Creating DNA reference database"
author: "Carol Flynn", "Gabriel A. Mendes"
format: html
editor: visual
editor_options:
  chunk_output_type: console
chunk_output_type: console
---

# Creating DNA reference database

## 1. Setup & Configuration

### Packages

-   **Load libraries.** Import all necessary packages for data manipulation, demultiplex, ASVs inference, taxonomy assigment and data visualization.

    ```{r}
    library(tidyverse)
    library(phyloseq)
    library(Biostrings)
    library(DECIPHER)
    library(writexl)
    library(readxl)
    library(BLASTr)
    library(phangorn)
    library(msa)
    library(ggtree)
    ```

### Directory Management

-   **Create directory structure.** Define project paths and create the directory structure required by the pipeline.

    ```{r}
    # Project folder path  
    prjct_path <- "/home/gabriel/projetos/roots_metabar"

    # Derived paths
    {
      env_path <- file.path(prjct_path, "environments")
      data_path <- file.path(prjct_path, "data")
      results_path <- file.path(prjct_path, "results")
      figs_path <- file.path(prjct_path, "figures")
      blast_path <- file.path(results_path, "blast")
      DB_path <- file.path(data_path, "reference_database/")
    }

    # Paths to be created
    paths <- c(prjct_path, env_path, data_path, results_path,
               figs_path, blast_path, DB_path)

    for (dir in paths) {
      if (dir.exists(dir)) {
    print(paste("The folder", dir, "already exists!"))
      } else {
    print(paste("Creating", dir, "!" ))
    dir.create(dir) 
      }
    }

    # Files in the project folder
    list.files(prjct_path)
    ```

-   **Load custom functions.** 

    ```{r}
    # Retrieve taxID custom function
    source("/home/gabriel/functions/retrieve_taxid.R")
    
    # Retrieve complete taxonomy custom function
    source("/home/gabriel/functions/retrieve_taxonomy.R")
    ```

### Raw data aquisition

-   **Upload raw metadata sheets**

    ```{r}
    # Metagen's samples output record
    metagen_output <-  
      read_csv(paste0(data_path, "/", "barcode/",
                      "samples_metagen-output.csv"))%>%
      tibble() %>% 
      select(-c(old_Tag, Tag))

    # Tanguro's sampling metadata 
    tanguro_leaf_metadata <- 
      read_csv(paste0(data_path,"/", "barcode/",
                      "leaf_sampling_tanguro_metadata.csv")) %>%
      tibble() %>% 
      select(-c(Tag))
    ```

## 2. Raw data processing

-   **Format tables.** Merge tissue reference metadata with the sample sequences

    ```{r}
    # Redux metadata table for previous data screening 
    tanguro_leaf_metadata_rdx <- tanguro_leaf_metadata %>% 
      select(c(Inventary_code, Popular_name, barcode_tag))

    # Associate Sanger reference sequences to correspondent metadata 
    database_output <- metagen_output %>% 
      left_join(tanguro_leaf_metadata, by = c("Popular_name", 
                                          "barcode_tag")) %>% 
      group_by(Popular_name) %>%
      fill(-Popular_name, .direction = "downup") %>%
      ungroup() %>% 
      mutate(`Species_clean` = sub(" .*", "", Species),
             Species_full = paste(Genus, Species_clean),
             Header = paste0(barcode_tag, "-", Genus, 
                             "_", Species_clean)) %>% 
      relocate(Header, Species_full)
    ```

-   **Retrieve full taxonomy.**

    ```{r}
    # Retrive taxIDs
    sequence_taxids <- retrieve_taxid(unique(database_output$Genus),
                                      ask=TRUE)

    # Retrieve complete taxonomy
    curated_taxonomy <- retrieve_taxonomy(sequence_taxids) %>%
      select(kingdom, phylum, class, order, family, genus, uid) %>%
      rename_with(stringr::str_to_title)
    
    # Join with reference and metadata
    database_output_full <- database_output %>% 
      left_join(curated_taxonomy, by = c("Genus", "Family")) %>% 
      relocate(Kingdom,
               Phylum,
               Class,
               Order,
               Family,
               Genus,
               Species_full,
               Species)
    ```

-   **Export full output table.**

    ```{r}
    writexl::write_xlsx(database_output_full,
                        path = file.path(results_path, 
                                         paste0("tanguro_DB", Sys.Date(), ".xlsx")))
    ```

## 3. Create DNA reference database

-   **Create DNAStringSet object.**

    ```{r}
    # Retrieve reference sequencers
    tanguro_DB <- DNAStringSet(database_output_tbl$Sequence)
    
    # Retrieve sequence's headers
    names(tanguro_DB) <- database_output_tbl$Header
    
    # Reorder based on names
    tanguro_DB <- tanguro_DB[order(names(tanguro_DB))]
    
    # Visualize DB
    BrowseSeqs(tanguro_DB)
    ```
    
-   **Edit DNAStringSet object.**

    ```{r}
    # Remove gaps
    tanguro_DB_nogaps <- RemoveGaps(tanguro_DB)
    BrowseSeqs(tanguro_DB_nogaps)
    
    # Orient nucleotides
    tanguro_DB_nogaps <- OrientNucleotides(tanguro_DB_nogaps)
    BrowseSeqs(tanguro_DB_nogaps)
    
    # Align sequences (didatic porpuses)
    tanguro_DB_aligned <- tanguro_DB_nogaps %>% AlignSeqs(refinements = 700,
                                                          iterations = 700,
                                                          verbose = TRUE)
    BrowseSeqs(tanguro_DB_aligned)
    ```
 
-   **Save DNA reference database fasta file.** 

    ```{r}
    writeXStringSet(tanguro_DB_nogaps, 
                    filepath = paste0(DB_path, "tanguro_DB_noGaps_", Sys.Date(), ".fasta"))
    ```

## 4. Create and plot DNA reference tree

-   **Create Neighbor-Joining Tree.**

    ```{r}
    # Align with ClustalW
    tanguro_DB_aligned <- msa(inputSeqs = tanguro_DB_nogaps,
                              method = "ClustalW",
                              substitutionMatrix = "clustalw",
                              type = "dna",
                              order = "input")
    
    # Convert alignment to phyDat
    phy_data <- as.phyDat(tanguro_DB_aligned, type = "DNA")
    
    #Calculate distance
    dist_phy <- dist.ml(phy_data, model = "JC69")
    
    # Create NJ tree
    tanguro_tree_nj <- NJ(dist_phy)
    ```

-   **Plot Neighbor-Joining Tree.**

    ```{r}
    tanguro_tree_plot <- ggtree(tanguro_tree_nj, ladderize = TRUE) %<+%
      database_output_tbl + 
      geom_tiplab(aes(label = Species_full, color = Family), offset = 0.01) +
      geom_treescale() +
      theme_tree2() +
      labs(title = "Tanguro's DNA Reference Database Tree")
    
    plot(tanguro_tree_plot)
    
    # Save tree plot
    ggsave(file = paste0(figs_path, "/", 
                                    PRJ, 
                                    "-primers_found_in_reads.pdf"),
           plot = primers_tile,
           device = "pdf",
           # width = ,
           # height = ,
           units = "cm",
           dpi = 600)
    ```